Verifying Android Applications using Java PathFinder

### 2012-10-02 Papers/paper.tex
Updated Paper

% paper.tex
% sample ACM SIG Proceedings document using LaTeX2e
% Author: Heila van der Merwe
% based upon LaTeX2.09 Guidelines, 9 June 1996
% Revisions:  17 July 2012
%
\documentclass{acm_proc_article-sp}
\begin{document}

@@

This paper describes the development of JPF-ANDROID, an Android application verification tool. JPF-ANDROID is built on Java Pathfinder, 
a Java model checking engine. JPF-ANDROID provides a simplified model of the Android framework on which an Android application can run. It then allows
the user to script input events to drive the application flow. JPF-ANDROID provides a way to detect common property violations such as deadlocks and runtime exceptions
in Android applications.
\end{abstract}

@@

The most common way to automate Android application testing is to run tests on the Dalvik VM on a
physical device/emulator. Testing frameworks such as the MonkeyRunner and Robotium make use of Android's built-in JUnit
framework~\cite{TestingAndroid} to run test suites on the device. Running tests on the Dalvik VM is slow as they have to be instrumented and
each test sequence has to be defined and executed sequentially. Other projects use alternative ways to automatically generate input by
manipulating the built in accessibility technologies~\cite{AccessibilityTech} or by re-implementing the Android
keyboard~\cite{KeyboardModel}. Mockito and Android Mock allow JUnit testing on the Dalvik VM by using mock Android classes. The advantage of
testing applications on the Dalvik VM is that we can physically emulate input events on the device. The disadvantage is that it is not
simple to automate this input emulation.

Another approach is to test Android applications using the Java Virtual Machine (JVM). There are different ways of implementing such
a framework. Android lint, for example, makes use of static analysis to identify common errors in applications.
Robolectric~\cite{robolectric}, a JUnit testing framework running on the JVM,  intercepts the loading of Android classes and then uses
shadow classes that model these classes.

@@

JPF-ANDROID makes use of JPF's extension mechanisms to model the Android application framework so that Android applications can run on the
JVM. It then extends JPF-AWT's input model to simulate user and system input to drive the application flow. One of the advantages of
extending JPF is that it has been rigorously tested and can successfully detect many common defects in Java applications. When we run
Android applications on JPF-ANDROID, common software errors are automatically detected.

@@

Android has one main application called the system process. The system process contains services responsible for performing the main
tasks of the system, including:
\vspace{-10pt}
\begin{description}
 \item [ActivityManager] manages the life-cycle and interaction of all the activities running on the system
 \item [WindowManager] allows applications to draw on the screen and forwards UI input to the application.
 \item [PackageManager] stores information on the application packages installed on the device
\end{description}
\vspace{-5pt}
All Android applications follow a single-threaded design in which the main thread of the application handles all application
events~\cite{AndroidDocs}. This structure is commonly used by many UI frameworks since it becomes too complex to make all UI classes thread
safe~\cite{SingleThread}. In Android, this main thread is called the \texttt{Looper} thread. The \texttt{Looper} has a message queue
attached to it containing all application events to be dispatched. UI and system events are scheduled on the main \texttt{Looper} by adding
them to the message queue. The \texttt{Looper} is responsible for continuously looping through these messages and handling them
appropriately. This could include updating a widget, loading a new activity or processing an \texttt{Intent} (see below).

@@

JPF-ANDROID focuses on verifying a single application with multiple application components and their interaction. The system service of
the Android OS is not part of the application process and runs in its own thread. To reduce scheduling possibilities, the entire system
service is not modelled but its necessary components are implemented as part of the application process. The following parts of the Android
framework is modelled:
\vspace{-10pt}
\begin{description}
 \item [ActivityManager]  manages the life-cycle of Activities and other application components.
 \item [ActivityThread]   the main entry-point to the application. It controls and manages the application components and their input.
 \item [Application components] including Activity, Service, Broadcast Receiver and Content Provider. 
 \item [Window and View structure] The view hierarchy is modelled including the widgets and the window classes.
 \item [Message queue] modelled to support script input.
\end{description}
\vspace{-5pt}
Lastly, as the application will not be communicating with outside processes, the \texttt{Intent} objects are modelled to exclude the
Binder IPC service.

@@

% \begin{figure}
% \centering
% \epsfig{file=newContact.eps, width=1in,}\\
% \caption{Contacts application window}
% \end{figure}
% \begin{figure}
% {\small
% {\sf
% ...\\
% \textbf{ANY} \{ \$cancelButton.onclick(), \$clearButton.onClick()  \}
% \$nameEdit.setText(``Mary'')\\
% \$createButton.onClick()\\
% ...
% }
% }
% \caption{Input script for ``New Contact'' window}
% \end{figure}
Let us take a Contacts Android application as an example. The application has two Activities: the \texttt{ListContactsActivity} and the
\texttt{NewContactActivity}. The \texttt{ListContactsActivity} contains a list of the stored contacts and a button to add a new contact that directs the
user to the \texttt{NewContactActivity}. The \texttt{NewContactActivity} contains two buttons: the clear button stays on the current Activity and clears its
fields and the create button starts the \texttt{ListContactsActivity}. Now let us look at Figure \ref{fig:contact} containing the input
sequence for the \texttt{NewContactActivity}.

@@

The problem occurs in the ANY structure. When the create button is clicked the application changes to the ListContactsActivity
and the following event is not valid any more. This drastically restricts the number of sequences that can be scripted in the file. To
address this issue, we included the use of sections in the input script (see Figure \ref{fig:contact2}). Each section groups the input
events of a specific Activity. Now, if the create button is pressed, the \texttt{ListContactsActivity} will be started and the events
specified in its section will start executing again.

@@

The addition of the sections in the input script lead to infinite event sequences. In the above example, an infinite loop occurs when the
create button is pressed on the \texttt{NewContactActivity} window. This will stop the execution of the \texttt{NewContactActivity} section and restart
the \texttt{ListContactsActivity} section's events. When the \texttt{ListContactsActivity}'s section is restarted, it again clicks on the create button
restarting the \texttt{NewContactActivity}. To address this issue, JPF-ANDROID keeps record of its current Activity. The scripting environment was
also adapted to store the current position in each visited Activity's script section. If a branch returns to a previously visited
Activity, its current position in its section is looked up and instead of restarting the section, it continues from its previous position.

@@

\begin{figure}
\frame{
\parbox{0.48\textwidth}{
{\small
{\sf 
\vspace*{2mm}
\hspace*{2mm}1. \textbf{SECTION} SimpleActivity  \{\\
\hspace*{2mm}2. \hspace*{5mm}\textbf{ANY} \{\,\$button[0-9].onClick()\,\}\\
\hspace*{2mm}3. \hspace*{5mm}\textbf{ANY} \{\,\$button<Plus|Minus|Mul|Div|More>.onClick()\,\}\\
\hspace*{2mm}4. \hspace*{5mm}\textbf{ANY} \{\,\$button[0-9].onClick()\,\}\\
\hspace*{2mm}5. \hspace*{5mm}\$buttonEquals.onClick()\\
\hspace*{2mm}6. \}\\
\hspace*{2mm}7. \textbf{SECTION} ScientificActivity  \{\\
\hspace*{2mm}8. \hspace*{5mm}\textbf{ANY} \{\,\$button<Sin|Cos|Tan>.onClick()\,\}\\
\hspace*{2mm}9. \}\\
}}}}
\vspace{-15pt}
\caption{Input script for Calculator application}
\label{fig:calc}
\vspace{-10pt}
\end{figure}

%what errors can occur
The calculator application contains two errors. When the user divides a value by zero, an uncaught
\texttt{ArithmeticException} is thrown by the application which causes Android to kill the application. Secondly, the application neglected
to attach the state information to the \texttt{Intent} that is passed to the next Activity. When the user switches to the other Activity a
\texttt{NullPointerException} is thrown when it tries to read the state information from the \texttt{Intent}.

To detect these errors we will use the test script in Figure~\ref{fig:calc}. The scripting environment interprets the script into the
following input sequences:
\begin{align*}
\text{<sequence>} &= \text{<simple\_seq>}\,|\,\text{<complex\_seq>}\\
\text{<simple\_seq>} &= \text{number},\,\text{simple\_op},\,\text{number},\,\text{equals}\\
\text{<complex\_seq>} &= \text{number},\,\text{complex\_op}\\
\text{simple\_op} &= \text{``+''}\,|\,\text{``-''}\,|\,\text{``}\times\text{''}\,|\,\text{``}\div\text{''}\\
\text{complex\_op} &= \text{``sin''}\,|\,\text{``cos''}\,|\,\text{``tan''}\\
\text{equals} &= \text{``=''} \\
\text{number} &= \text{``0''\,|\,``1''\,|\,``2''\,|\,``3''\,|\,``4''\,|\,``5''\,|\,``6''\,|\,``7''\,|\,``8''\,|\,``9''}
\end{align*}
Each of the ten number buttons are pressed followed by a simple operation and another number or a complex operation. Although it is possible
to input numbers consisting of multiple digits including floating point numbers, for illustration purposes this was left out of the input script.
JPF is automatically configured to detect thrown exceptions and to stop execution. When JPF-ANDROID was run on the application it firstly
detected the \texttt{ArithmenticException} due to the division by zero:

@@

Another extension that will be added to JPF-ANDROID is coverage testing. JPF has many coverage extensions available so we will be adapting
one these extensions to work on Android applications~\cite{coverage}. Coverage testing is especially important to find errors related to the Android Activity
life cycle such as the null pointer exception mentioned above.

@@

% \vspace*{5pt}
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\scriptsize
\bibliography{paper}  % sigproc.bib is the name of the Bibliography in this case

>>>>

% paper.tex
% sample ACM SIG Proceedings document using LaTeX2e
% Author: Heila van der Merwe
% based upon LaTeX2.09 Guidelines, 9 June 1996
% Revisions:  17 July 2012
%
\documentclass{acm_proc_article-sp}
\usepackage{etoolbox}
\makeatletter
\patchcmd{\maketitle}{\@copyrightspace}{}{}{}
\makeatother
% \usepackage{flushend}

\begin{document}

@@

This paper describes the development of JPF-ANDROID, an Android application verification tool. JPF-ANDROID is built on Java Pathfinder, 
a Java model checking engine. JPF-ANDROID provides a simplified model of the Android framework on which an Android application can run. It then allows
the user to script input events to drive the \linebreak application flow. JPF-ANDROID provides a way to detect common property violations such as deadlocks and runtime exceptions
in Android applications.
\end{abstract}

@@

The most common way to automate Android application testing is to run tests on the Dalvik VM on a
physical device/emulator. Testing frameworks such as the MonkeyRunner~\cite{TestingAndroid} and Robotium~\cite{robotium} make use of Android's built-in JUnit
framework~\cite{TestingAndroid} to run test suites on the device. Running tests on the Dalvik VM is slow as they have to be instrumented and
each test sequence has to be defined and executed sequentially. Other projects use alternative ways to automatically generate input by
manipulating the built in accessibility technologies~\cite{AccessibilityTech}, by re-implementing the Android
keyboard~\cite{KeyboardModel} or by instrumenting the Dalvik VM to generate logs that are analysed to detect common errors~\cite{hu}. Mockito~\cite{mockito} and Android Mock~\cite{androidMock} allow JUnit testing on the Dalvik VM by using mock Android classes.  The advantage of
testing applications on the Dalvik VM is that we can physically emulate input events on the device. The disadvantage is that it is not
simple to automate this input emulation.

Another approach is to test Android applications using the Java Virtual Machine (JVM). There are different ways of implementing such
a framework. Android Lint, for example, makes use of static analysis to identify common errors in applications.
Robolectric~\cite{robolectric}, a JUnit testing framework running on the JVM,  intercepts the loading of Android classes and then uses
shadow classes that model these classes.

@@

JPF-ANDROID makes use of JPF's extension mechanisms to model the Android application framework so that Android applications can run on the
JVM. It then extends JPF-AWT's input model to simulate user and system input to drive the application flow. One of the advantages of
extending JPF is that it has been rigorously tested and can successfully detect many common defects in Java applications. When Android applications
run on JPF using JPF-ANDROID, common software errors are automatically detected.

@@

Android has one main application called the system process. The system process contains services responsible for performing the main
tasks of the system, including:
\vspace{-10pt}
\begin{description}
 \item [ActivityManager] manages the life-cycle and interaction of all the activities running on the system
 \item [WindowManager] allows applications to draw on the screen and forwards UI input to the application.
 \item [PackageManager] stores information on the application \linebreak packages installed on the device
\end{description}
\vspace{-5pt}
All Android applications follow a single-threaded design in which the main thread of the application handles all application
events~\cite{AndroidDocs}. This structure is commonly used by many UI frameworks since it becomes too complex to make all UI classes thread
safe~\cite{SingleThread}. In Android, this main thread is called the \texttt{Looper} thread. The \texttt{Looper} has a message queue
attached to it containing all application events to be dispatched. UI and system events are scheduled on the \texttt{Looper} by adding
them to the message queue. The \texttt{Looper} is responsible for continuously looping through these messages and handling them
appropriately. This could include updating a widget, loading a new activity or processing an \texttt{Intent} (see below).

@@

JPF-ANDROID focuses on verifying a single application with multiple application components and their interaction. The system service of
the Android OS is not part of the application process and runs in its own thread. To reduce scheduling possibilities, the entire system
service is not modelled but its necessary components are implemented as part of the application process. The following parts of the Android
framework is modelled:
\vspace{-10pt}
\begin{description}
 \item [ActivityManager]  manages the life-cycle of Activities and other application components.
 \item [ActivityThread]   the main entry-point to the application. It controls and manages the application components and their input.
 \item [Application components] including Activity, Service, \linebreak Broadcast Receiver and Content Provider. 
 \item [Window and View structure] The view hierarchy is modelled including the widgets and the window classes.
 \item [Message queue] modelled to support script input.
\end{description}
\vspace{-5pt}
Lastly, as the application will not be communicating with outside processes, the \texttt{Intent} objects are modelled to exclude the
Binder IPC service.

@@

% \begin{figure}
% \centering
% \epsfig{file=newContact.eps, width=1in,}\\
% \caption{Contacts application window}
% \end{figure}
% \begin{figure}
% {\small
% {\sf
% ...\\
% \textbf{ANY} \{ \$cancelButton.onclick(), \$clearButton.onClick()  \}
% \$nameEdit.setText(``Mary'')\\
% \$createButton.onClick()\\
% ...
% }
% }
% \caption{Input script for ``New Contact'' window}
% \end{figure}
Let us take a Contacts Android application as an example. The application has two Activities: the \texttt{ListContactsActivity} and the
\texttt{NewContactActivity}. The \texttt{ListContactsActivity} contains a list of the stored contacts and a button to add a new contact that directs the
user to the \texttt{NewContactActivity}. The \texttt{NewContactActivity} contains two buttons: the clear button stays on the current Activity and clears its
fields and the create button starts the \texttt{ListContactsActivity}. Figure \ref{fig:contact} contains the\texttt{NewContactActivity}'s events.

@@

The problem occurs in the ANY structure. When the create button is clicked the application changes to the \texttt{ListContactsActivity}
and the following event is not valid any more. This drastically restricts the number of sequences that can be scripted in the file. To
address this issue, we included the use of sections in the input script (see Figure \ref{fig:contact2}). Each section groups the input
events of a specific Activity. Now, if the create button is pressed, the \texttt{ListContactsActivity} will be started and the events
specified in its section will start executing again.

@@

The addition of the sections in the input script lead to infinite event sequences. In the above example, an infinite loop occurs when the
create button is pressed on the \texttt{NewContactActivity} window. This will stop execution of the \texttt{NewContactActivity} section and restart
the \texttt{ListContactsActivity} section's events. When the \texttt{ListContactsActivity}'s section is restarted, it again clicks on the create button
restarting the \texttt{NewContactActivity}. To address this issue, JPF-ANDROID keeps record of its current Activity. The scripting environment was
also adapted to store the current position in each visited Activity's section. If a branch returns to a previously visited
Activity, its current position in its section is looked up and instead of restarting the section, it continues from its previous position.

@@

\begin{figure}
\frame{
\parbox{0.48\textwidth}{
{\small
{\sf 
\vspace*{2mm}
\hspace*{2mm}1. \textbf{SECTION} SimpleActivity  \{\\
\hspace*{2mm}2. \hspace*{5mm}\textbf{ANY} \{\,\$button[0-9].onClick()\,\}\\
\hspace*{2mm}3. \hspace*{5mm}\textbf{ANY} \{\,\$button<Plus|Minus|Mul|Div|More>.onClick()\,\}\\
\hspace*{2mm}4. \hspace*{5mm}\textbf{ANY} \{\,\$button[0-9].onClick()\,\}\\
\hspace*{2mm}5. \hspace*{5mm}\$buttonEquals.onClick()\\
\hspace*{2mm}6. \}\\
\hspace*{2mm}7. \\
\hspace*{2mm}8. \textbf{SECTION} ScientificActivity  \{\\
\hspace*{2mm}9. \hspace*{5mm}\textbf{ANY} \{\,\$button<Sin|Cos|Tan>.onClick()\,\}\\
\hspace*{2mm}10. \}\\
}}}}
\vspace{-15pt}
\caption{Input script for Calculator application}
\label{fig:calc}
\vspace{-10pt}
\end{figure}

%what errors can occur
The calculator application contains two errors. When the user divides a value by zero, an uncaught
\texttt{ArithmeticException} is thrown by the application which causes Android to kill the application. Secondly, the application neglected
to attach the state information to the \texttt{Intent} that is passed to the next Activity. When the user switches to the other Activity a
\texttt{NullPointerException} is thrown when it tries to read the state information from the \texttt{Intent}. To detect these errors we will use the test script in Figure~\ref{fig:calc}. The scripting environment interprets the script into the
following input sequences:
\begin{align*}
\text{<sequence>} &= \text{<simple\_seq>}\,|\,\text{<complex\_seq>}\\
\text{<simple\_seq>} &= \text{number},\,\text{simple\_op},\,\text{number},\,\text{equals}\\
\text{<complex\_seq>} &= \text{number},\,\text{complex\_op}\\
\text{simple\_op} &= \text{``+''}\,|\,\text{``-''}\,|\,\text{``}\times\text{''}\,|\,\text{``}\div\text{''}\\
\text{complex\_op} &= \text{``sin''}\,|\,\text{``cos''}\,|\,\text{``tan''}\\
\text{equals} &= \text{``=''} \\
\text{number} &= \text{``0''\,|\,``1''\,|\,``2''\,|\,``3''\,|\,``4''\,|\,``5''\,|\,``6''\,|\,``7''\,|\,``8''\,|\,``9''}
\end{align*}
Each of the ten number buttons are pressed followed by a simple operation and another number or a complex operation. Although it is possible
to input numbers consisting of multiple digits including floating point numbers, for illustration purposes this was left out of the input script.
JPF is automatically configured to detect thrown exceptions and to stop execution. When JPF-ANDROID was run on the application it firstly
detected the \texttt{ArithmenticException} due to the division by zero:

@@

Another extension that will be added to JPF-ANDROID is coverage testing. JPF has many coverage extensions avail-
\balancecolumns
able so we will be adapting
one these extensions to work on Android applications~\cite{coverage}. Coverage testing is especially important to find errors related to the Android Activity
life cycle such as the null pointer exception mentioned above.

@@

% \vspace*{5pt}
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
% \bibliography{sigproc}
\scriptsize
\bibliography{paper}  % sigproc.bib is the name of the Bibliography in this case
